(With Extensions)The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-13 rational_ext.tex
created an _ext version for more notes

\title{The Two Dualities of Computation: Negative and Fractional Types}
\authorinfo{Roshan P. James}
           {Indiana University}
           {rpjames@indiana.edu}
\authorinfo{Amr Sabry}
           {Indiana University}
           {sabry@indiana.edu}
\maketitle

@@

This section introduces the syntax and graphical languages for
negatives and fractionals.  The general outline is as follows:
\footnote{We refer the reader to Selinger's excellent survey
  article of monoidal
  categories~\cite{springerlink:10.1007/978-3-642-12821-94} for the precise
  definitions.}
\begin{itemize}
\item As established in our earlier work~\cite{rc2011,infeffects}, the
  underlying categorical semantics of our core reversible language
  {{langRev}} is based on \emph{two} distinct symmetric monoidal structures,
  one with {{+}} as the monoidal tensor, and one with {{*}} as the monoidal
  tensor.
\item We extend each underlying symmetric monoidal structure to a
  \emph{compact closed} structure by adding a dual for each object and two
  special morphisms traditionally called {{eta}} and {{eps}}. 
\end{itemize}
The extended language is referred to as {{langRevEE}}.  After
presenting the extension at the syntactic level, we discuss the
categorical semantics informally via the graphical language. The
operational semantics is presented in Sec.~\ref{sec:rat}.

@@

>>>>

\title{(With Extensions)The Two Dualities of Computation: Negative and Fractional Types}
\authorinfo{Roshan P. James}
           {Indiana University}
           {rpjames@indiana.edu}
\authorinfo{Amr Sabry}
           {Indiana University}
           {sabry@indiana.edu}
\maketitle

@@

This section introduces the syntax and graphical languages for
negative types and fractionals.The general outline is as follows:
\footnote{We refer the reader to Selinger's excellent survey
  article of monoidal
  categories~\cite{springerlink:10.1007/978-3-642-12821-94} for the precise
  definitions.}
\begin{itemize}
\item As established in our earlier work~\cite{rc2011,infeffects}, the
  underlying categorical semantics of our core reversible language
  {{langRev}} is based on \emph{two} distinct symmetric monoidal structures,
  one with {{+}} as the monoidal tensor, and one with {{*}} as the monoidal
  tensor.
\item We extend each underlying symmetric monoidal structure to a
  \emph{compact closed} structure by adding a dual for each object and two
  special morphisms traditionally called {{eta}} and {{eps}}. 
\end{itemize}
After presenting the extension at the syntactic level, we discuss the
categorical semantics informally via the graphical language. The operational
semantics is presented in Sec.~\ref{sec:rat}.

@@

It is useful to contrast the categorical structure of {{langRevEE}},
which has two distinct \emph{compact closed} structures, to the
well-known and closed related \emph{*-autonomous} categories that form
a model for linear logic. The distnguishing feature that seperates
*-autonomous structures from compact closed structures is that the
dualizing action in *-autonomous categories does not preserve the
tensor. In {{langRev}}, as we will see , we can construct the
isomorphism {{-(b1+b2)<-> (-b1)+(-b2)}}, i.e. the tensor {{+}} has
been preserved by the negation.  Whereas in linear logic we have that
$(a\oplus b)^\bot$ is $a^\bot \parr b^\bot$, i.e. the connective
$\oplus$ is not preserved.

In linear logic terms `additive pairs' and `multiplicative sums' are
used to indicate this notion that some connectives has both additive
and multiplicative properties. The usual intuitive interpretations for
these involve suggestions like one has a dollar to spend and the both
cigarettes and soda are available for purchase, but the fact that you
have only a dollar to spend affords you only one of these. The
intuitions talk about a sort of dependacy underlying the fact that
both cigarettes and soda are available but you can have only one.

In {{langRevEE}}, operators {{eta}} and {{eps}} capture a similar
notion.The pair of values {{v}} and {{1/v}} created by {{eta*}} are
not fully independant in the way one would think of normal pairs. The
value of one, affects the value of the other. Similarly, the {{v}} and
{{-v}} created by {{eta+}} do really present a strict choice of
{{left}} or {{right}} in the sense that usual sums do, but instead can
have a \emph{both} {{left}} and {{right}} values at different points
during the computation. However, in our opinion, the philisophical
difference the additiveness of the additive pairs and the
multiplicativesness of the multiplicative pair aren't a property of
the connective between the values, but a property intrinsic to the
values themselves. Hence {{v}} and {{1/v}} share secretely a
relationship that their combinting {{*}} constructor does not and need
not know about.
